---
title: "HDAC inhibitor screen analysis"
author: "F. Heigwer"
date: "06/02/2021"
output:
  pdf_document: 
    keep_tex: yes
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
#Lets define some themes
#This we do as a quality of life step to give all figures that we produce a common look and feel.

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(FitAR)
library(corrr)
library(ggrastr)
library(ggrepel)
library(broom)
library(patchwork)

theme_b110 <- function(){
  theme_classic() +
  theme(
    axis.text = element_text(size = 10), 
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12,hjust = 0.5,face = "bold"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = "bottom"
    )
}

sgi_blue    = '#5087C8'
sgi_yellow1 = '#F2EE35'
sgi_yellow2 = '#FED98E'
b110_grey   = '#808080'
b110_grey_light   = '#909090'
b110_transparent_black = scales::alpha('#000000',0.5)
google_red = '#dd4b39'
google_green = '#0F9D58'
google_yellow = '#F4B400'
google_blue = '#4285F4'

```

## Overview

This vignette is used to analyze a CRISPR screen performed by treating SW480 cells using 8 nM panobinostat in 0.1 % DMSO and 0.1 % DMSO as a control sample. It was screened in duplicates per treatment. Raw counts per sgRNA are the basis of the analysis.

All plots were used as given for the main figure panels. The compiled vignette pdf serves as an overview.

## Data input

```{r data_input, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}

treat1<-read_delim(file = "../raw_data_counts/counts_panobinostat_screen/P1_S7_L001_R1_001.txt",delim = "\t")
treat2<-read_delim(file = "../raw_data_counts/counts_panobinostat_screen/P2_S8_L001_R1_001.txt",delim = "\t")

dmso1<-read_delim(file = "../raw_data_counts/counts_panobinostat_screen/DMSO1_S1_L001_R1_001.txt",delim = "\t")
dmso2<-read_delim(file = "../raw_data_counts/counts_panobinostat_screen/DMSO2_S2_L001_R1_001.txt",delim = "\t")

mydata<-treat1 %>% 
  left_join(treat2,by="sgRNA") %>% 
  left_join(dmso1,by="sgRNA") %>% 
  left_join(dmso2,by="sgRNA") %>%
  extract(sgRNA,c("sgRNA","gene_id"),"((\\w+)_.*)") 

names(mydata)<-c("sgRNA","gene_id","treat_1","treat_2","dmso_1","dmso_2")

mydata %<>%
  gather(treatment,count,-sgRNA,-gene_id) %>%
  extract(treatment,c("treatment","replicate"),"(\\w+)_(\\d)")

anno<-read_delim("../meta_data/gene_names.txt",delim="\t")

```

## Lets normalize the data

```{r Normalization, echo=TRUE}

mydata %<>% 
  mutate(count=if_else(count>10,count,NA_real_)) %>%    #filter low read counts
  group_by(treatment,replicate) %>% #group by treatment and replicate
  mutate(normval=asinh(count/median(count,na.rm = T)))

```

## Lets do some data QC

### sgRNA-level replicate correlation

```{r QC1, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

mydata %>% 
  select(-count) %>%
  spread(replicate,normval) %>%
  split(.$treatment) %>%
  lapply(function(x){x %>% ungroup() %>% select(`1`,`2`) %>% correlate() %>% rearrange() %>% shave() %>% fashion()})
  
# on gene level we get a replicate read count correlation of 0.94 and 0.97 for DMSO and Panobinostat treated samples respectively
```

### gene-level replicate correlation

```{r QC2, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

mydata %>% 
  group_by(treatment,gene_id,replicate) %>%
  summarise(mean_normval=mean(normval,na.rm=T)) %>%
  spread(replicate,mean_normval) %>%
  split(.$treatment) %>%
  lapply(function(x){x %>% ungroup() %>% select(`1`,`2`) %>% correlate() %>% rearrange() %>% shave() %>% fashion()})
  
# on gene level we get a replicate read count correlation of 0.94 and 0.97 for DMSO and Panobinostat treated samples respectively
```

### graphics

```{r QC3, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}


p1 <- mydata %>% 
  select(-count) %>%
  spread(replicate,normval) %>%
  ggplot(aes(x=`1`,y=`2`)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(~treatment,scales = "free") +
    theme_b110() +
    ggtitle("sgRNA-level read counts")

#ggsave("../graphics/normalized_readcount_replicates_sgRNA_Panobinostat.pdf",p1,width = 8,height = 8)

# from this analysis we get a replicate correlation of 0.88 and 0.91 for the DMSO and Panobinostat treated samples respectively

p2 <- mydata %>% 
  select(-count) %>%
  group_by(treatment,gene_id,replicate) %>%
  summarise(mean_normval=mean(normval,na.rm=T)) %>%
  spread(replicate,mean_normval) %>%
  ggplot(aes(x=`1`,y=`2`)) +
    geom_point() +
    geom_smooth(method = "lm")+
    facet_wrap(~treatment,scales = "free") +
    theme_b110()+
    ggtitle("gene-level read counts")

#ggsave("../graphics/normalized_readcount_replicates_genes_Panobinostat.pdf",p2,width = 8,height = 8)

print(p1+p2)
```

QC looks very nice and good so lets calls some foldchanges and hits thereof

## Hit-calling 

```{r differential_folchanges, echo=FALSE, message=FALSE, warning=FALSE}

#we average replicates and plot treated versus untreated samples
#sgRNA level
p1 <- mydata %>% 
  group_by(sgRNA,gene_id,treatment) %>%
  summarise(value=mean(normval,na.rm=T)) %>%
  spread(treatment,value) %>%
  ggplot(aes(x=dmso,y=treat)) +
    geom_point() +
    theme_b110() +
    ggtitle("sgRNA-level differentials")

#ggsave("../graphics/readcount_dmso_vs_vorinostat_sgRNA_Panobinostat.pdf",p1,width = 8,height = 8)

#gene level
p2 <- mydata %>% 
  group_by(gene_id,treatment) %>%
  summarise(value=mean(normval,na.rm=T)) %>%
  spread(treatment,value) %>%
  ggplot(aes(x=dmso,y=treat)) +
    geom_point() +
    geom_abline(slope = 1,intercept = c(0,0)) +
    theme_classic()+
    theme_b110()+
    ggtitle("gene-level differentials")

#ggsave("../graphics/readcount_dmso_vs_vorinostat_genes_Panobinostat.pdf",p2,width = 8,height = 8)

print(p1+p2)

```

### Statistical analysis

Here we use pairwise wilcoxon rank-sum test between to fold changes of the individual sgRNAs for each gene and the individual negative control sgRNA to determine if the foldchange of the gene represented by independent sgRNAs is significantly different from the non-targeting controls.

```{r hitcalling, echo=FALSE, message=FALSE, warning=FALSE}
#lets call some foldchanges and hits
tmp<-mydata %>% 
  group_by(sgRNA,gene_id,treatment) %>%
  summarise(value=mean(normval,na.rm=T)) %>%
  spread(treatment,value) %>%
  ungroup() %>%
  mutate(logfc=log2(treat/dmso)) 

stats<- tmp %>%
  group_by(gene_id) %>%
  do(
   wilcox.test(.$logfc,tmp$logfc[tmp$gene_id=="random"],alternative = "t") %>% tidy()
  ) %>%
  ungroup() %>%
  mutate(fdr=p.adjust(p.value,method="BH"))

final_sg<- tmp %>% left_join(stats) %>% mutate(significant=if_else(fdr<0.1,"significant","non-significant"))

final_gene<-
  final_sg %>%
  group_by(gene_id) %>%
  summarise(dmso=mean(dmso,na.rm = T),treat=mean(treat,na.rm = T),logfc=mean(logfc,na.rm = T),p.value=unique(p.value),fdr=unique(fdr),significant=unique(significant))

#gene level
p1 <- final_gene %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  ggplot(aes(x=dmso,y=treat,color=significant,label=labels)) +
    geom_point() +
    geom_abline(slope = 1,intercept = c(0,0)) +
    theme_classic() +
    geom_label_repel()+
    scale_color_manual(values = c(b110_grey_light,google_blue))+
    ggtitle("gene-level differentials marked")


#ggsave("../graphics/readcount_dmso_vs_vorinostat_genes_stats_Panobinostat.pdf",p1,width = 8,height = 8)

#gene level
p2 <- final_gene %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  ggplot(aes(x=logfc,y=-log10(p.value),color=significant,label=labels)) +
    geom_point() +
    theme_classic() +
    geom_label_repel() +
    scale_color_manual(values = c(b110_grey_light,google_blue))+
    ggtitle("gene-level volcano")

#ggsave("../graphics/volcano_stat_colored_Panobinostat.pdf",p2,width = 8,height = 8)

print(p1+p2)

final_gene %>% 
  left_join(anno) %>% 
  write_delim("../results/GeneLevelData_Panobinostat.tsv",delim = "\t")


p1 <- final_sg %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  filter(!is.na(labels)) %>%
  mutate(cols=factor(sign(logfc))) %>%
  ggplot(aes(x=reorder(sgRNA,logfc),y=logfc,fill=cols)) +
    geom_bar(stat = "identity",position = "dodge")  +
  facet_wrap(~gene_symbol,scales = "free") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
    scale_fill_manual(values = c(google_blue,google_red))+
    ggtitle("candidate waterfall plots")


#ggsave("../graphics/EHMT_sgRNAs_colored_Panobinostat.pdf",p1,width = 8,height = 8)

print(p1)

final_sg %>% 
  left_join(anno) %>% 
  mutate(labels=if_else(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B"),gene_symbol,NA_character_)) %>%
  filter(!is.na(labels)) %>%
  mutate(cols=factor(sign(logfc))) %>% 
  write_delim("../results/EHMT1_EHMT2_sgRNALevelData_Panobinostat.tsv",delim = "\t")

final_sg %>% 
  left_join(anno) %>% 
  filter(gene_id == "random") %>%
  write_delim("../results/random_sgRNALevelData_Panobinostat.tsv",delim = "\t")

p2 <- final_sg %>% 
  left_join(anno) %>% 
  filter(gene_symbol %in% c("EHMT1","EHMT2","CARM1","TADA2B") | gene_id %in% c("random")) %>%
  mutate(labels = if_else(is.na(gene_symbol),gene_id,gene_symbol)) %>%
  mutate(cols = factor(sign(logfc))) %>%
  select(labels,sgRNA,logfc,cols) %>%
  drop_na() %>%
  ggplot(aes(x = reorder(sgRNA,logfc), y = logfc, fill = cols)) +
    geom_bar(stat = "identity",position = "dodge")  +
  facet_wrap(~labels,scales = "free_x") +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
    scale_fill_manual(values = c(google_blue,google_red))+
    ggtitle("control waterfall plots")


#ggsave("../graphics/EHMT_random_sgRNAs_colored_Panobinostat.pdf",p2,width = 8,height = 8)

print(p2)
```

# Session info

```{r echo=FALSE, message=FALSE, warning=FALSE}

sessionInfo()

#writeLines(capture.output(sessionInfo()), paste("../results/SessionInfo_Panobinostat.txt",sep = ""))
```
